The Chemistry Development Kit (CDK): atom typing, rendering, molecular formulas, and substructure searching

### 2016-10-09 article.tex
More tweaks and updated the builder section based on the TODOs given before

Aromaticity has seen many definitions in the past and for
cheminformatics it frequently is algorithmically defined. The outcome
of an aromaticity calculation depends on a number of atom type
features and heuristics, which are often ambiguously defined in the
published literature. Based on this information several different
algorithmic definitions of aromaticity can be defined. Older CDK
versions had various models implemented and the code was scattered
throughout the library, resulting in an inconsistent approach to
calling methods to compute aromaticity and a significant maintenance
burden.  This was unified in the current version, resulting in three
models, of which two are based on the CDK atom typer. The difference
between these two models is how contributions from exocyclic double
bonds are handled.

@@

\subsubsection*{New Builders}

Originally, the CDK was developed as a shared library between
JChemPaint~\cite{krause2000jchempaint} and
Jmol~\cite{Willighagen2007jmol,Hanson2010}.  JChemPaint used a MVC
approach with an event-passing mechanism to update the view when the
model was changed. This can cause an cascade of change events being
passed around. This was not always a desirable feature, especially for
non-UI code. To address this, interfaces were introduced allowing
multiple implementations of the core interfaces.
% TODO: check if the interfaces have previously been defined
The \texttt{IChemObjectBuilders} play an important role here, allowing
implementations of the interfaces to be instantiated without the need
of explicitly referencing those implementations.\todo{The motivation
  for IChemObjectBuilders is not well defined.}

However, the CDK 1.0 and 1.2 implementations of the \texttt{IChemObjectBuilder} had one method for
each constructor, resulting in a very large interface. Moreover, the API changed whenever
a new class was introduced, and existing methods changed when constructors were updated.
To simplify the API, the new \texttt{IChemObjectBuilder} collapsed all methods returning new
implementations into a single method, which takes as a first parameter the class of the
interface that is wished to be constructed. All further parameters are passed as
parameters to the class constructor.

For example, to construct a new atom from its element symbol, one
would now write:\todo{might be good to display code using older version}

\vspace{0.2cm}
\begin{verbatim}
IChemObjectBuilder builder; // previously defined
IAtom atom = builder.newInstance(IAtom.class, "C");
\end{verbatim}
\vspace{0.2cm}

The CDK library is now written against the interfaces\todo{this may
  not mean to a non-expert Java coder}, and when new instances are
needed, these builders are used. This allows to run a certain
CDK-based application with a specific builder, aimed at a particular
use case. The original implementation is available via the
\texttt{DefaultChemObjectBuilder}, which creates classes that pass
around change events, as in the older CDK versions.  However, a
\texttt{SilentChemObjectBuilder} is available, that does not support
change events, resulting in speedups of 10\%-20\%.  The third builder
is the \texttt{DataDebugChemObjectBuilder} which generates debug
information for all changes to the content of the data classes. This
can be useful for debugging and other forms of code inspection.

\subsubsection*{Molecular fingerprints}
Molecular fingerprints are another area that seen significant
development. Traditionally fingerprints were represented using the
\texttt{BitSet} class from the Java library. While using this class
allowed the use of pre-existing methods to manipulate bit strings, it
keeps a vector of bits in memory. The solution was excellent for
hashed, relatively small fingerprints, \textit{e.g.}, 1024 bits,
\textit{i.e.}, 10 bits indexing space. However implementing a
fingerprint designed to avoid collisions with, \textit{e.g.} a 32 bit
indexing space using this approach would be highly
memory-inefficient. To allow for multiple fingerprint representation
implementations a bit fingerprint interface was introduced\todo{is
  there any more information about this?}. Also,
although fingerprints traditionally are bit vectors a count
fingerprint was also introduced making fingerprints based on integer
vectors supported in CDK as well. The fingerprints currently existing
in CDK are listed in Table~3\ref{tab:fingerprints}.

>>>>

Aromaticity has seen many definitions in the past and for
cheminformatics it frequently is algorithmically defined. The outcome
of an aromaticity calculation depends on a number of atom type
features and heuristics, which are often ambiguously defined in the
published literature. Based on the information used, several different
algorithmic definitions of aromaticity can be defined. Older CDK
versions had various aromaticity models implemented but the code was scattered
throughout the library, resulting in an inconsistent API
to compute aromaticity and a significant maintenance
burden.  The API was unified in the current version, resulting in three
models, of which two are based on the CDK atom typer. The difference
between these two models is how contributions from exocyclic double
bonds are handled.

@@

\subsubsection*{New Object Builders}

Originally, the CDK was developed as a shared library between
JChemPaint~\cite{krause2000jchempaint} and
Jmol~\cite{Willighagen2007jmol,Hanson2010}.  JChemPaint used a MVC
approach with an event-passing mechanism to update the view when the
model was changed. This can cause an cascade of change events being
passed around. This was not always a desirable feature, especially for
non-UI code. To address this, interfaces were introduced allowing
multiple implementations of the core interfaces. With much code of the CDK
library no longer based on the original data model, a builder is needed to
create objects of that data model, such as an implementation of the IAtom.
The new \texttt{IChemObjectBuilders} allow implementations to be created, allowing
implementations of the interfaces to be instantiated without the need
of explicitly referencing those implementations. This way, any algorithm
implementation in the CDK can use any of the data model interface
implementations.

The CDK 1.0 and 1.2 implementations of the \texttt{IChemObjectBuilder} had,
however, one method for each data object constructor, resulting in a very large
interface. Moreover, this interface API had to be updated each time a new class
was introduced, and when existing methods changed and constructors were updated.
To simplify the API, the new \texttt{IChemObjectBuilder} collapses all methods
into a single method, which takes as a first parameter the class of the
interface that is to be constructed. All further parameters are passed as
parameters to the class constructor.

For example, to construct a new atom from its element symbol, one
would write previously:

\vspace{0.2cm}
\begin{verbatim}
IChemObjectBuilder builder = ...;
IAtom atom = builder.newAtom("C");
\end{verbatim}
\vspace{0.2cm}

With the new builder, the code looks like:

\vspace{0.2cm}
\begin{verbatim}
IChemObjectBuilder builder = ...;
IAtom atom = builder.newInstance(IAtom.class, "C");
\end{verbatim}
\vspace{0.2cm}

The CDK library is now mostly written that it no longer depends on a specific
implementation of the \texttt{IChemObjectBuilder}, allowing the user of the
CDK to select a builder suitable to their software. Therefore, if software
depends on event passing, then the \texttt{DefaultChemObjectBuilder} can be
used:

\vspace{0.2cm}
\begin{verbatim}
IChemObjectBuilder builder = DefaultChemObjectBuilder.getInstance();
\end{verbatim}
\vspace{0.2cm}

However, if events are not needed, a \texttt{SilentChemObjectBuilder} is
available, resulting in a typical speedup of 10\%-20\%.  The third builder
is the \texttt{DataDebugChemObjectBuilder} which generates debug
information for all changes to the content of the data classes. This
can be useful for debugging and other forms of code inspection.

\subsubsection*{Molecular Fingerprints}
Molecular fingerprints are another area that seen significant
development. Traditionally fingerprints were represented using the
\texttt{BitSet} class from the Java library. While using this class
allowed the use of pre-existing methods to manipulate bit strings, it
keeps a vector of bits in memory. The solution was excellent for
hashed, relatively small fingerprints, \textit{e.g.}, 1024 bits,
\textit{i.e.}, 10 bits indexing space. However implementing a
fingerprint designed to avoid collisions with, \textit{e.g.} a 32 bit
indexing space using this approach would be highly
memory-inefficient. To allow for multiple fingerprint representation
implementations a bit fingerprint interface was introduced\todo{is
  there any more information about this?}. Also,
although fingerprints traditionally are bit vectors a count
fingerprint was also introduced making fingerprints based on integer
vectors supported in CDK as well. The fingerprints currently existing
in CDK are listed in Table~3\ref{tab:fingerprints}.
